---
title: "team-Peru-penguins-lab"
format: html
---

```{r}
#| label: setup
#| message: false
library(tidyverse)
library(ggbeeswarm)
dim(penguins)
```

# Data Preparation

```{r}
#| label: setup-dataset
#| message: false
dim(penguins)
```

## Size of the Dataset
The data frame has 344 rows and 8 columns.

## Preliminary Exploration

### Using head()

```{r}
#| label: head-of-dataframe
#| message: true
#| include: true
#| echo: true
head(penguins)
```

The head() output shows the first six rows of the penguins dataset. The first penguins are from the Adelie species on Torgersen island, and the dataset includes measurements such as bill length, bill depth, flipper length, body mass, sex, and year.

### Using tail()

```{r}
#| label: tail-of-dataframe
#| message: true
#| include: true
#| echo: true
tail(penguins)
```
The tail() output shows the last six rows, which are Chinstrap penguins from Dream island. This suggests that the dataset is arranged by species and/or island rather than randomly mixed.

### Using glimpse()
```{r}
#| label: glimpse-of-dataframe
#| message: true
#| include: true
#| echo: true
glimpse(penguins)
```

The glimpse() output shows that the dataset has 344 rows and 8 columns. Some variables are categorical, such as species, island, and sex, while others are numeric, such as bill_len, bill_dep, flipper_len, and body_mass. There are also missing values, for example in row 4 where bill length, bill depth, flipper length, body mass, and sex are missing.

## Data Cleaning
```{r}
#| label: penguins_bill-tibble
#| message: true
#| echo: true
#| include: true
penguins_bill <- penguins |>
  select(species, bill_len, bill_dep) |>
  drop_na()

nrow(penguins_bill)
```

There are 342 rows left in the data frame as 2 rows were removed due to having NA data.

## Verifier Task
```{r}
#| label: verifier-row-check
#| message: true
#| echo: true
#| include: true
rows_removed <- penguins |>
  filter(
    is.na(bill_len) | is.na(bill_dep) | is.na(species)
  ) |>
  nrow()

rows_removed
```

# Distribution of Bill Length

## Ungrouped Bill-Length Histogram
```{r}
#| label: bill-length-histogram
#| message: true
#| echo: true
#| include: true
#| fig-width: 7
#| fig-height: 5
ggplot(penguins_bill, aes(x = bill_len)) +
  geom_histogram(binwidth = 2, color = "white") +
  geom_rug(
    aes(y = 0),
    sides = "b",
    alpha = 0.35,
    position = position_jitter(width = 0.2, height = 0)
  ) +
  labs(
    title = "Distribution of Penguin Bill Length",
    x = "Bill length (mm)",
    y = "Number of penguins",
    caption = "Data source: Gorman et al. (2014)"
  )
```

### Verifier Task
```{r}
#| label: verifier-bill-length
#| message: true
#| echo: true
#| include: true
bill_range <- range(penguins_bill$bill_len)
bill_range
```
The shortest bill is 32.1 mm and the longest is 59.6 mm. The histogram covers this full range — nothing is cut off at either end.

## Bin Width

### Bin Width 4
```{r}
#| label: bill-length-histogram-width-4
#| message: true
#| echo: true
#| include: true
#| fig-width: 4
#| fig-height: 5
ggplot(penguins_bill, aes(x = bill_len)) +
  geom_histogram(binwidth = 2, color = "white") +
  geom_rug(
    aes(y = 0),
    sides = "b",
    alpha = 0.35,
    position = position_jitter(width = 0.2, height = 0)
  ) +
  labs(
    title = "Distribution of Penguin Bill Length",
    x = "Bill length (mm)",
    y = "Number of penguins",
    caption = "Data source: Gorman et al. (2014)"
  )
```

### Bin width 10
```{r}
#| label: bill-length-histogram-width-10
#| message: true
#| echo: true
#| include: true
#| fig-width: 10
#| fig-height: 5
ggplot(penguins_bill, aes(x = bill_len)) +
  geom_histogram(binwidth = 2, color = "white") +
  geom_rug(
    aes(y = 0),
    sides = "b",
    alpha = 0.35,
    position = position_jitter(width = 0.2, height = 0)
  ) +
  labs(
    title = "Distribution of Penguin Bill Length",
    x = "Bill length (mm)",
    y = "Number of penguins",
    caption = "Data source: Gorman et al. (2014)"
  )
```

### Discussion
We compared histograms using 4 mm, 7 mm, and 10 mm bin widths. The 4 mm bin width shows more detail and still reveals that the bill-length distribution is not unimodal, but it has more small fluctuations between neighbouring bins. The 10 mm bin width is too coarse and wide, making the distribution look overly simplified. We prefer the 7 mm bin width for the report because it gives a cleaner and more presentable overall shape while still showing that the distribution contains more than one cluster. This makes it easier for readers to understand the main pattern without being distracted by too much bin-level noise.

## Histogram Interpretation
The bill-length distribution is not unimodal because the dataset combines multiple penguin species. Each species has a different typical bill length, so pooling them together creates several clusters instead of one single peak. This suggests that species is an important grouping variable and should not be ignored when interpreting bill length.

## Raw Point Plot by Species
```{r}
#| label: raw-point-bill-length-by-species
#| message: true
#| echo: true
#| include: true
#| fig-width: 7
#| fig-height: 5
ggplot(penguins_bill, aes(x = bill_len, y = species)) +
  geom_point() +
  labs(
    title = "Bill Length by Species Using Raw Points",
    x = "Bill length (mm)",
    y = "Species",
    caption = "Data source: Gorman et al. (2014)"
  ) +
  theme_minimal()
```

### Observation
Using geom_point() with the default settings, the points for each species appear on three horizontal rows because species is a categorical variable. Many points overlap or are very close together, especially where many penguins have similar bill lengths. As a result, the plot shows the approximate range of bill lengths for each species, but it does not clearly show the density of points along each row. This makes it difficult to tell where most penguins are concentrated.

## Jitter Plot
```{r}
#| label: jitter-bill-length-by-species
#| message: true
#| echo: true
#| include: true
#| fig-width: 7
#| fig-height: 5
set.seed(3107)

ggplot(penguins_bill, aes(x = bill_len, y = species)) +
  geom_jitter(
    width = 0,
    height = 0.2,
    alpha = 0.5
  ) +
  labs(
    title = "Bill Length by Species Using Jitter",
    x = "Bill length (mm)",
    y = "Species",
    caption = "Data source: Gorman et al. (2014)"
  ) +
  theme_minimal()
```

### Observation
The jitter plot makes the density of bill lengths easier to see compared with the raw point plot. Adelie penguins are mostly concentrated at shorter bill lengths, roughly around 35–43 mm. Gentoo and Chinstrap penguins generally have longer bill lengths, with many values around the mid-40s to low-50s. However, Gentoo and Chinstrap overlap quite a lot, so bill length alone does not clearly separate these two species. This pattern was harder to see in the raw point plot because many points overlapped along each species row.

## Beeswarm Plot
```{r}
#| label: beeswarm-bill-length-by-species
#| message: true
#| echo: true
#| include: true
#| fig-width: 7
#| fig-height: 5
ggplot(penguins_bill, aes(x = bill_len, y = species)) +
  geom_beeswarm(alpha = 0.6) +
  labs(
    title = "Bill Length by Species Using Beeswarm",
    x = "Bill length (mm)",
    y = "Species",
    caption = "Data source: Gorman et al. (2014)"
  ) +
  theme_minimal()
```

### Observation
The beeswarm plot shows the bill length distribution for each species while reducing overlap between points. Compared with the jitter plot, the points are arranged more systematically, so dense regions are easier to identify. Adelie penguins are mostly concentrated at shorter bill lengths, roughly around 35–43 mm. Gentoo penguins are mostly concentrated around the mid-40s to low-50s, while Chinstrap penguins are also mostly in the mid-40s to low-50s. This confirms that bill length separates Adelie from the other two species fairly well, but Gentoo and Chinstrap still overlap, so bill length alone cannot fully distinguish all three species.

## Jitter Versus Beeswarm

### Advantage
One advantage of the beeswarm plot is that it shows the density of bill lengths more clearly than the jitter plot. In the beeswarm plot, the points are arranged systematically, so it is easier to see where many penguins are concentrated, especially for Adelie around the shorter bill lengths and Gentoo around the mid-40s to low-50s. 

### Disadvantage
One disadvantage is that the beeswarm plot can look more visually packed and less familiar to readers than a jitter plot. The jitter plot is simpler to understand, but its random spreading makes the pattern look less organized.

## Faceted-Bil-Length Hitogram
```{r}
#| label: faceted-bill-length-histogram
#| message: true
#| echo: true
#| include: true
#| fig-width: 8
#| fig-height: 5
set.seed(3107)

overall_mean_bill_len <- mean(penguins_bill$bill_len)

ggplot(penguins_bill, aes(x = bill_len)) +
  geom_histogram(
    binwidth = 7,
    boundary = 0,
    fill = "grey75",
    color = "white"
  ) +
  geom_rug(
    aes(y = 0),
    sides = "b",
    alpha = 0.18,
    position = position_jitter(width = 0.08, height = 0)
  ) +
  geom_vline(
    xintercept = overall_mean_bill_len,
    linetype = "dashed"
  ) +
  facet_wrap(~species) +
  labs(
    title = "Bill Length Distribution by Species",
    x = "Bill length (mm)",
    y = "Number of penguins",
    caption = "Data source: Gorman et al. (2014)"
  ) +
  theme_minimal()
```

### Observation
The faceted histogram shows that the overall bill-length distribution is made up of different species-specific distributions. Adelie penguins are mostly concentrated at shorter bill lengths, with most observations below the overall mean line. Chinstrap and Gentoo penguins generally have longer bill lengths, with many observations around the mid-40s to low-50s. The dashed vertical line shows the overall mean bill length across all penguins and provides a common reference point across the three facets. This helps explain why the ungrouped histogram was not unimodal: it combined species with different bill-length distributions.

## Violin-With-Box Plot
```{r}
#| label: violin-box-bill-length
#| message: true
#| echo: true
#| include: true
#| fig-width: 7
#| fig-height: 5
ggplot(penguins_bill, aes(x = bill_len, y = species)) +
  geom_violin(
    fill = "grey85",
    color = "grey40"
  ) +
  geom_boxplot(
    width = 0.15,
    fill = "white",
    outlier.alpha = 0.5
  ) +
  labs(
    title = "Bill Length by Species Using Violin and Box Plot",
    x = "Bill length (mm)",
    y = "Species",
    caption = "Data source: Gorman et al. (2014)"
  ) +
  theme_minimal()
```

### Observation
The violin-with-box plot shows that Adelie penguins have noticeably shorter bill lengths than Chinstrap and Gentoo penguins. The Adelie distribution is concentrated mostly around the high-30s to low-40s mm range, while Chinstrap and Gentoo are shifted further right, mostly around the mid-40s to low-50s mm range. The box plots show that the median bill length for Adelie is lower than the medians for Chinstrap and Gentoo. The violin shapes also show that Chinstrap and Gentoo have overlapping bill-length distributions, meaning bill length alone does not clearly separate those two species. Gentoo also has a long right tail with a very high bill-length value near 60 mm, suggesting a possible extreme value or outlier.

## Why Combine Violin and Box Plots
The violin plot shows the full distribution shape, making it easier to see where bill lengths are more concentrated for each species. However, it does not clearly show summary statistics such as the median and interquartile range. The box plot provides these summary statistics, but it hides the detailed distribution shape. Combining the two gives both the density pattern and the key numerical summaries in one visualization.

# Distribution of Bill Depth

## Ungrouped Bill-Depth Histogram
```{r}
#| label: bill-depth-histogram
#| message: true
#| echo: true
#| include: true
#| fig-width: 7
#| fig-height: 5
set.seed(3107)

ggplot(penguins_bill, aes(x = bill_dep)) +
  geom_histogram(
    binwidth = 0.5,
    boundary = 0,
    fill = "grey75",
    color = "white"
  ) +
  geom_rug(
    aes(y = 0),
    sides = "b",
    alpha = 0.18,
    position = position_jitter(width = 0.03, height = 0)
  ) +
  labs(
    title = "Distribution of Penguin Bill Depth",
    x = "Bill depth (mm)",
    y = "Number of penguins",
    caption = "Data source: Gorman et al. (2014)"
  ) +
  theme_minimal()
```

### How does the distribution of bill depth compare to that of bill length
The ungrouped bill-depth histogram shows that bill depth ranges roughly from about 13 mm to 21 mm. The distribution is not perfectly unimodal because there are several visible clusters, especially around 14–15 mm and 17–19 mm. Most penguins appear to have bill depths around 17–19 mm, where the tallest bars are located. Compared with the bill-length histogram, bill depth has a narrower numerical range, but it still shows signs that the data may contain different species groups. The rug marks along the bottom show the individual penguin observations and confirm that many values are concentrated in the middle of the range.

## Faceted Bill-Depth Histogram
```{r}
#| label: faceted-bill-depth-histogram
#| message: true
#| echo: true
#| include: true
#| fig-width: 8
#| fig-height: 5
set.seed(3107)

overall_mean_bill_dep <- mean(penguins_bill$bill_dep)

ggplot(penguins_bill, aes(x = bill_dep)) +
  geom_histogram(
    binwidth = 0.5,
    boundary = 0,
    fill = "grey75",
    color = "white"
  ) +
  geom_rug(
    aes(y = 0),
    sides = "b",
    alpha = 0.18,
    position = position_jitter(width = 0.03, height = 0)
  ) +
  geom_vline(
    xintercept = overall_mean_bill_dep,
    linetype = "dashed"
  ) +
  facet_wrap(~species) +
  labs(
    title = "Bill Depth Distribution by Species",
    x = "Bill depth (mm)",
    y = "Number of penguins",
    caption = "Data source: Gorman et al. (2014)"
  ) +
  theme_minimal()
```

### How does the distribution of bill depth differ across species?
The faceted histogram shows that bill depth differs clearly across species. Gentoo penguins have the shallowest bill depths, with most values concentrated below the overall mean line and mostly around 14–16 mm. Adelie penguins have deeper bills, with most values around 17–19.5 mm and many observations above the overall mean. Chinstrap penguins also have relatively deep bills, mostly around 17–20 mm, and they overlap more with Adelie than with Gentoo. This suggests that bill depth is especially useful for separating Gentoo from Adelie and Chinstrap, but it does not separate Adelie and Chinstrap as clearly.

## Comparison of Bill-Length and Bill-Depth Distributions
Bill length and bill depth separate the penguin species in different ways. From the bill-length visualizations in Section 3.2(e–j), Adelie penguins are the easiest to separate because they generally have shorter bill lengths than Chinstrap and Gentoo penguins. However, Chinstrap and Gentoo overlap quite a lot in bill length, so bill length alone does not clearly separate those two species.

From the bill-depth histogram in Section 3.3(b), Gentoo penguins are the easiest to separate because they generally have shallower bill depths than Adelie and Chinstrap penguins. Adelie and Chinstrap have deeper bill depths and overlap more with each other. Therefore, the two measurements do not separate exactly the same species: bill length mainly separates Adelie from the others, while bill depth mainly separates Gentoo from the others.

## Verifier Task
```{r}
#| label: verify-species-bill-means
penguins_bill |>
  summarize(
    mean_bill_len = mean(bill_len),
    mean_bill_dep = mean(bill_dep),
    .by = species
  )
```
The numbers back up what the plots showed:

- Bill length: Adelie has the shortest average (≈ 38 mm), Chinstrap is in the
  middle (≈ 49 mm), and Gentoo is the longest (≈ 47 mm). Chinstrap and Gentoo are
  close to each other in length, which explains why they overlap in the histograms.
- Bill depth: Gentoo is clearly the shallowest (≈ 15 mm), while Adelie (≈ 18 mm)
  and Chinstrap (≈ 18 mm) are almost identical — confirming that depth separates
  Gentoo from the rest but cannot split Adelie from Chinstrap.


# Birate View: Bill Depth Versus Bill Length

## Ungrouped Scatter Plot
```{R}
#| label: ungrouped-scatter-plot
#| message: true
#| echo: true
#| include: true
#| fig-width: 7
#| fig-height: 5

ggplot(penguins_bill, aes(x = bill_len, y = bill_dep)) +
  geom_point(alpha = 0.6) +
  geom_smooth(method = "lm", se = TRUE) +
  labs(
    title = "Bill Depth versus Bill Length",
    x = "Bill length (mm)",
    y = "Bill depth (mm)",
    caption = "Data source: Gorman et al. (2014)"
  ) +
  theme_minimal()
```

### Observation
The ungrouped scatter plot shows the relationship between bill length and bill depth for all penguins combined. Each point represents one penguin, and the blue regression line shows the pooled linear trend across all species. The line slopes downward, suggesting that when species is ignored, penguins with longer bill lengths appear to have lower bill depths. However, the points seem to form several clusters, so the pooled trend may be hiding species-level patterns.

## Interpretation Of The Pooled Hrend
If species information were unavailable, a reader would infer a negative relationship between bill length and bill depth. In other words, the pooled trend suggests that penguins with longer bills tend to have shallower bills. The confidence band around the regression line is fairly narrow and does not suggest a flat relationship, so the pooled trend appears statistically meaningful. However, this interpretation may be misleading because the plot combines different species into one group.

## Exact Overplotting
```{r}
#| label: exact-overplotting
#| message: true
#| echo: true
#| include: true
overlap_counts <- penguins_bill |>
  count(bill_len, bill_dep)

hidden_points <- overlap_counts |>
  summarize(hidden = sum(n - 1)) |>
  pull(hidden)

max_stack <- overlap_counts |>
  summarize(max_stack = max(n)) |>
  pull(max_stack)

hidden_points
max_stack
```

### Discussion
There are 4 points hidden by exact overlap. The maximum number of penguins stacked at a single (bill_len, bill_dep) coordinate is 2. This means exact duplicate coordinates exist, but the amount of exact overlap is quite small compared with the full cleaned dataset.

Exact overlap is not the only cause of overplotting. Even if two points do not have exactly the same bill length and bill depth, they can still be very close together visually, especially in dense regions of the scatter plot. This can make the plot look crowded and can make it difficult to judge where most observations are concentrated.

### Verifier Task
```{r}
#| label: overplot-distinct-verify
#| message: true
#| echo: true
#| include: true
visible_points <- penguins_bill |> distinct(bill_len, bill_dep) |> nrow()
hidden_distinct <- nrow(penguins_bill) - visible_points

hidden_distinct
```

The verifier calculation using distinct() also gives 4 hidden points, confirming the result from count().

## Mitigation Of Overplotting
```{r}
#| label: jittered-scatter-plot
#| message: true
#| echo: true
#| include: true
#| fig-width: 7
#| fig-height: 5
set.seed(3107)

ggplot(penguins_bill, aes(x = bill_len, y = bill_dep)) +
  geom_jitter(
    width = 0.15,
    height = 0.15,
    alpha = 0.5
  ) +
  geom_smooth(method = "lm", se = TRUE) +
  labs(
    title = "Jittered Scatter Plot of Bill Depth versus Bill Length",
    x = "Bill length (mm)",
    y = "Bill depth (mm)",
    caption = "Data source: Gorman et al. (2014)"
  ) +
  theme_minimal()
```

```{r}
#| label: count-symbol-scatter-plot
#| fig-width: 7
#| fig-height: 5
ggplot(penguins_bill, aes(x = bill_len, y = bill_dep)) +
  geom_count(alpha = 0.7) +
  scale_size_area(max_size = 8) +
  geom_smooth(method = "lm", se = TRUE) +
  labs(
    title = "Count Plot of Bill Depth versus Bill Length",
    x = "Bill length (mm)",
    y = "Bill depth (mm)",
    size = "Number of penguins",
    caption = "Data source: Gorman et al. (2014)"
  ) +
  theme_minimal()
```

## Assessment of Scatter Plot Variants
We would submit the jittered scatter plot to a journal referee. The exact-overlap check showed that only 4 points were hidden by identical coordinates, with a maximum stack size of 2, so exact duplicate points are not a serious issue in this dataset. The count-symbol plot shows exact overlaps, but the large circles make the plot look more crowded and harder to interpret. The jittered scatter plot makes dense regions easier to see while still preserving the overall negative pooled trend, so it is the clearer option for communicating the pattern.

## Grouped Scatter Plot
```{r}
#| label: grouped-jittered-scatter-plot
#| message: true
#| echo: true
#| include: true
#| fig-width: 7
#| fig-height: 5
set.seed(3107)

ggplot(penguins_bill, aes(x = bill_len, y = bill_dep, color = species)) +
  geom_jitter(
    width = 0.15,
    height = 0.15,
    alpha = 0.6
  ) +
  geom_smooth(method = "lm", se = TRUE) +
  labs(
    title = "Bill Depth versus Bill Length by Species",
    x = "Bill length (mm)",
    y = "Bill depth (mm)",
    color = "Species",
    caption = "Data source: Gorman et al. (2014)"
  ) +
  theme_minimal()
```

### Observation
Mapping species to colour reveals that the penguins form three separate species clusters. Adelie penguins generally have shorter bill lengths and deeper bills, Gentoo penguins have shallower bill depths, and Chinstrap penguins tend to have longer and deeper bills. The regression lines for all three species slope upward, showing a positive relationship between bill length and bill depth within each species. This pattern was hidden in the ungrouped scatter plot, where the pooled regression line sloped downward. The apparent reversal between the pooled trend and the species-specific trends is an example of Simpson’s paradox.

### Verifier Task
```{r}
#| label: correlation-check
overall_cor <- cor(penguins_bill$bill_len, penguins_bill$bill_dep)
round(overall_cor, 3)

penguins_bill |>
  summarise(
    r = round(cor(bill_len, bill_dep), 3),
    .by = species
  )
```
The overall correlation is -0.235 — negative, matching the downward-sloping pooled line. Within each species every correlation is positive, matching the upward-sloping per-species lines in the grouped plot. This confirms Simpson’s Paradox in the numbers: the sign of the relationship literally flips depending on whether you account for species.

# Summarize The Insights Gained
Bill length and bill depth can help distinguish the three Pygoscelis penguin species, especially when both measurements are considered together. The bill-length plots show that Adelie penguins generally have shorter bills than Chinstrap and Gentoo penguins. However, bill length alone does not fully separate Chinstrap and Gentoo because their distributions overlap. The bill-depth plots show a different pattern: Gentoo penguins generally have shallower bills than Adelie and Chinstrap penguins. This means bill length and bill depth separate different species and provide complementary information. When species is ignored in the ungrouped scatter plot, the pooled regression line suggests a negative relationship between bill length and bill depth. However, the grouped scatter plot shows positive relationships within each species, so the misleading reversal is an example of Simpson’s paradox.

## Verifier Task
Our findings are consistent with what Horst, Hill and Gorman (2022) report. They introduced the Palmer Penguins dataset specifically as a teaching example that illustrates Simpson's Paradox, and they highlight that the negative pooled correlation between bill length and bill depth (~−0.23) reverses to a positive correlation within each species — exactly what we observed. They also note that the three species form reasonably distinct clusters in the two-dimensional bill space, which matches our conclusion that both measurements are needed together to approach full separation.